What We Eval in the Shadows

### 2021-08-09 paper/main.tex
3

\maketitle

@@

The corpus is assembled from three sources: the \emph{Base} libraries bundled
with the language implementation, packages hosted on \emph{CRAN}, and scripts
from \emph{Kaggle}. Each of these sources represents a different part of
the user base of the language and thus, potentially, different usage patterns.

\mypara{Base.} The \BasePackages libraries are bundled with R and widely used as
they provide arithmetics, statistics, and operating system functionalities. Some
\BaseFunsWithEvals functions contain \BaseEvalCallSites calls to \eval. The
functions perform such key tasks such as package loading, there is thus hardly
any R program that does not invoke them. These libraries have been written by
experts with a deep understanding of the language and its environment.

\mypara{CRAN.} This dataset consists of packages hosted on the Comprehensive R
Archive Network ({\small \url{cran.r-project.org}}, aka CRAN). There are three
sources of runnable code in a package: unit \emph{tests} for individual
functions, code \emph{examples} automatically extracted from the documentation,
and long-form use-cases called \emph{vignettes}. Package also often have sample
input files. All of these runnable code snippets can be extract into independent
files. We downloaded \CranPackages packages and extracted \CranRunnableScripts
scripts, or \CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in
examples, \CranRunnableCodeVignettesRnd in vignettes and
\CranRunnableCodeTestsRnd in tests). CRAN packages are authored by experienced
developers.

\mypara{Kaggle.} This dataset contains \KaggleUnique programs from the Kaggle
website ({\small \url{kaggle.com}}), an online platform that allows users to
submit and compete to solve problems. Kaggle contains both scripts and notebooks
as well as input files, but there is some duplication (we used SHA-1 hashes to
remove \KaggleDuplicates duplicates). The dataset has 665K lines of R code and
only \KaggleWithEvals calls \eval. Kaggle scripts are written by end-users with
variable levels of expertise.

@@

 \begin{figure}[ht!]
   \centering
   \scalebox{0.85}{
     \begin{tikzpicture}
       \definecolor{red}{HTML}{F8CECC}
       \definecolor{darkred}{HTML}{B85450}
       \definecolor{yellow}{HTML}{FFF2CC}
       \definecolor{darkyellow}{HTML}{D6B656}
       \definecolor{blue}{HTML}{DAE8FC}
       \definecolor{darkblue}{HTML}{6C8EBF}
       \definecolor{green}{HTML}{D5E8D4}
       \definecolor{darkgreen}{HTML}{82B366}

       \newcommand{\nodesep}[0]{0.035 \textwidth}
       \newcommand{\textsep}[0]{0.005 \textwidth}

       \newcommand{\nodename}[1]{\begin{tabular}{c}#1\end{tabular}}
       \newcommand{\nodedesc}[1]{\begin{tabular}{c}#1\end{tabular}}

       \tikzstyle{block}     = [rectangle, rounded corners, minimum width=0.15 \textwidth, minimum height=60pt]
       \tikzstyle{connector} = [line width=0.25mm, ->]

       \node [block, fill = red, draw = darkred, very thick] (download) {
         \nodename{
           \vspace{1mm}\textbf{Download}\vspace{1mm}\\
           15K Packages\\
           8K Kaggle Kernels\\
           219 Kaggle Datasets
         }
       };
       \node [below = \textsep of download]   (downloaddesc)   {\nodedesc{12 Hours}};

       \node [block, right = \nodesep of download, fill = yellow, draw = darkyellow, very thick] (extract) {
         \nodename{
           \vspace{1mm}\textbf{Extract}\vspace{1mm}\\
           240K package programs\\%, 4.6M LOC\\
           74 Kaggle kernels\\%, 13K LOC\\
           \newline
         }
       };
       \node [below = \textsep of extract]   (extractdesc)   {\nodedesc{2 Hours}};

       \node [block, right = \nodesep of extract, fill = blue, draw = darkblue, very thick] (trace) {
         \nodename{
           \vspace{1mm}\textbf{Trace}\vspace{1mm}\\
           49M eval calls \\
           14M side-effects\\
           \newline
         }
       };
       \node [below = \textsep of trace]     (tracedesc)     {\nodedesc{1 Day 12 Hours}};

       \node [block, right = \nodesep of trace, fill = green, draw = darkgreen, very thick] (analyze) {
         \nodename{
           \vspace{1mm}\textbf{Analyze}\vspace{1mm}\\
           2.8M unique eval calls\\
           22\% packages use eval\\
           \newline
         }
       };
       \node [below = \textsep of analyze]    (analyzedesc)    {\nodedesc{2 Hours}};

       \draw [connector] (download)   edge (extract);
       \draw [connector] (extract)    edge (trace);
       \draw [connector] (trace)      edge (analyze);
     \end{tikzpicture}
   }
   \vspace{-2mm}\caption{Pipeline}\label{fig:pipeline}
 \end{figure}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage Metrics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Base}\label{sec:usage-base}

We recorded \baseAllcalls \eval calls in the \BasePackages base libraries. This
comes from running 10\% randomly selected programs from the
\CranRunnableScriptsRnd extracted programs from CRAN packages, covering
\baseTriggeredevalpct of the \BaseEvalCallSites sites. Most of the calls
(\baseEvalsratio) are to the \c{eval} function, and most arguments
(\baseCodepercent) are expressions. However, unlike CRAN, the majority
(\baseLanguagepercent) of the arguments are language objects. The median
argument size is \baseMedianszeval, which is more than for CRAN and the maximum
size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$)
amount to \baseSmalleventspct of calls. A single site in the \c{match.arg}
function is responsible for \baseTopFuncPercent of all recorded calls. This
function provides a convenient way for argument verification using partial
matching. It is heavily relied on in R functions that use string arguments to
parameterize function behavior, common practice in R. For example, in a body of
\c{center<-function(x,type=c("mean","median","trimmed"))} one can get the
value of \c{type} parameter using \c{type<-match.arg(type)}. A call
\c{center(x,"trim")} will match \c{type} to \c{"trimmed"}. The challenge
dealing with Base is that every program uses it. It can thus generate extreme
amounts of data, but the data is quite predictable. From now on, we only mention
Base when it surprises.

>>>>

\maketitle
\renewcommand{\shortauthors}{A. Goel, P. Donat-Bouillud, F. Křikava, C. Kirsch and J. Vitek}

@@

We distinguish three sources for calls to \eval, those originating from
\emph{Base} libraries bundled with R, packages hosted on \emph{CRAN}, and
end-user scripts from \emph{Kaggle}. We consider base libraries as part of the
language implementation and do not include them in our analysis.

\mypara{Base.} The \BasePackages base libraries provide arithmetics, statistics,
and operating system functionalities. Base has \BaseFunsWithEvals functions with
\BaseEvalCallSites calls to \eval. These functions perform key tasks such as
package loading, there is thus hardly any R program that does not invoke them.

\mypara{CRAN.} Packages hosted on the Comprehensive R Archive Network ({\small
  \url{cran.r-project.org}}, aka CRAN) have three sources of runnable: unit
\emph{tests} for individual functions, code \emph{examples} embedded in the
documentation, and use-cases called \emph{vignettes}. All runnable snippets can
be extracted into independent files. We download \CranPackages packages and
extract \CranRunnableScripts scripts with \CranRunnableCode lines of code
(\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
vignettes and \CranRunnableCodeTestsRnd in tests). CRAN packages are authored by
experienced developers.

\mypara{Kaggle.} Kaggle ({\small \url{kaggle.com}}) is an online platform that
allows users to submit and compete to solve problems. We download \KaggleUnique
unique scripts and notebooks with their input files (\KaggleDuplicates
duplicates were removed using SHA-1 hashes); 665K lines of R code making only
\KaggleWithEvals calls \eval. Authors of these scripts have highly variable
levels of expertise.

\mypara{Discussion.} When we started this project, our goal was to contrast the
usage of \eval by package developers and end-users, but the Kaggle dataset have
so few calls to \eval that there is little that can be said about those. The
remainder of the paper focuses on CRAN with a few observations about the other
sources when relevant.

@@

 \begin{figure}[ht!]
   \centering
   \scalebox{0.85}{
     \begin{tikzpicture}
       \definecolor{red}{HTML}{F8CECC}
       \definecolor{darkred}{HTML}{B85450}
       \definecolor{yellow}{HTML}{FFF2CC}
       \definecolor{darkyellow}{HTML}{D6B656}
       \definecolor{blue}{HTML}{DAE8FC}
       \definecolor{darkblue}{HTML}{6C8EBF}
       \definecolor{green}{HTML}{D5E8D4}
       \definecolor{darkgreen}{HTML}{82B366}
       \newcommand{\nodesep}[0]{0.035 \textwidth}
       \newcommand{\textsep}[0]{0.005 \textwidth}
       \newcommand{\nodename}[1]{\begin{tabular}{c}#1\end{tabular}}
       \newcommand{\nodedesc}[1]{\begin{tabular}{c}#1\end{tabular}}
       \tikzstyle{block}     = [rectangle, rounded corners, minimum width=0.15 \textwidth, minimum height=60pt]
       \tikzstyle{connector} = [line width=0.25mm, ->]

       \node [block, fill = red, draw = darkred, very thick] (download) {
         \nodename{
           \vspace{1mm}\textbf{Download}\vspace{1mm}\\
           15K Packages\\
           8K Kaggle Kernels\\
           219 Kaggle Datasets
         }
       };
       \node [below = \textsep of download]   (downloaddesc)   {\nodedesc{12 Hours}};
       \node [block, right = \nodesep of download, fill = yellow, draw = darkyellow, very thick] (extract) {
         \nodename{
           \vspace{1mm}\textbf{Extract}\vspace{1mm}\\
           240K package programs\\%, 4.6M LOC\\
           74 Kaggle kernels\\%, 13K LOC\\
           \newline
         }
       };
       \node [below = \textsep of extract]   (extractdesc)   {\nodedesc{2 Hours}};
       \node [block, right = \nodesep of extract, fill = blue, draw = darkblue, very thick] (trace) {
         \nodename{
           \vspace{1mm}\textbf{Trace}\vspace{1mm}\\
           49M eval calls \\
           14M side-effects\\
           \newline
         }
       };
       \node [below = \textsep of trace]     (tracedesc)     {\nodedesc{1 Day 12 Hours}};
       \node [block, right = \nodesep of trace, fill = green, draw = darkgreen, very thick] (analyze) {
         \nodename{
           \vspace{1mm}\textbf{Analyze}\vspace{1mm}\\
           2.8M unique eval calls\\
           22\% packages use eval\\
           \newline
         }
       };
       \node [below = \textsep of analyze]    (analyzedesc)    {\nodedesc{2 Hours}};
       \draw [connector] (download)   edge (extract);
       \draw [connector] (extract)    edge (trace);
       \draw [connector] (trace)      edge (analyze);
     \end{tikzpicture}
   }
   \vspace{-2mm}\caption{Pipeline}\label{fig:pipeline}
 \end{figure}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Measuring Eval across Datasets}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Base}\label{sec:usage-base}
We recorded \baseAllcalls \eval calls in the base libraries. This
comes from running 10\% randomly selected programs from the
\CranRunnableScriptsRnd extracted programs from CRAN packages, covering
\baseTriggeredevalpct of the \BaseEvalCallSites sites. Most of the calls
(\baseEvalsratio) are to the \c{eval} function, and most arguments
(\baseCodepercent) are expressions. However, unlike CRAN, the majority
(\baseLanguagepercent) of the arguments are language objects. The median
argument size is \baseMedianszeval, which is more than for CRAN and the maximum
size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$)
amount to \baseSmalleventspct of calls. A single site in the \c{match.arg}
function is responsible for \baseTopFuncPercent of all recorded calls. This
function provides a convenient way for argument verification using partial
matching. It is heavily relied on in R functions that use string arguments to
parameterize function behavior, common practice in R. For example, in a body of
\c{center<-function(x,type=c("mean","median","trimmed"))} one can get the
value of \c{type} parameter using \c{type<-match.arg(type)}. A call
\c{center(x,"trim")} will match \c{type} to \c{"trimmed"}. The challenge
dealing with Base is that every program uses it. It can thus generate extreme
amounts of data, but the data is quite predictable. From now on, we only mention
Base when it surprises.
